---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Maya Lis"
date: 9/10/26
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---


## Setup

```{r}
library(tidyverse)
library(tidycensus)
```


---

## The question

**Every number** the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.


**In this analysis, I look at [Median age] across counties in [Florida], and
ask which communities we measure most and least reliably.**

---

## Choosing what to measure


```{r}
my_state <- "FL"        # two-letter abbreviation, in quotes
my_variable <- "B01002_001"     # the variable code, in quotes
```


my_state <- "FL"        
my_variable <- "B01002_001" 

county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```

**[There are 67 rows which matches the number of counties in Florida.]**

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)

county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```


**[Lafeytte County has the largest margin of error relative to its estimate. It has the largest margin of error relative to its estimate at 6.22%]**

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 2  ~ "High confidence",
    moe_pct < 4 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```


**[I changed the reliability thresholds to 2% and 4% since the original 5% and 10% thresholds would have placed most counties in the high-confidence category. Of Florida's 67 counties, 59 have high-confidence estimates, 5 have moderate-confidence estimates, and 3 have low-confidence estimates.]**

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 2  ~ "High confidence",
    moe_pct < 4  ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```


**[High-confidence counties have an average population of 369,666. Low-confidence counties have an average population of 9,336. According to this pattern, smaller counties tend to have less reliable estimates of median age.This means policymakers should be careful about making decisions based on the median-age estimates for small counties since there's more uncertainty in the estimates.]**



**[We know the least about Lafayette county, Liberty county, and Glades county. Lafayette County has the highest relative margin of error at 6.22%, followed by Liberty County at 5.37%. The higher margin of error means that there is more uncertainty around the median-age estimates for these counties than for most other counties in Florida.Given the pattern displayed in the analysis, these counties have smaller populations and therefore larger margin of errors. The ACS uses a sample of households rather than collecting information from every person.Smaller counties have fewer people and so have smaller samples. With fewer observations, it is harder to produce precise estimates. This further causes a larger margin of error. This can definitely create problems for equitable funding throughout the state.If a state agency used median age to allocate funding it would create problems if they assumed each counties estimate is equally reliable. Major government service like healthcare and senior services can be based off of the age counts in counties.If the estimate for a small county is inaccurate, the county could receive too much or too little funding. This would most likley impact young and older populations who often require greater care. It's safe to assume smaller counties are vulnerable to having their needs overlooked and so policymakers should consider MOE when using ACS data.]**

